A Cost-based Approach for Fast Intrusion Detection

### 2017-06-05 tex/paper_v1.tex
final final version >.<

The aim of real time Intrusion Detection Systems(IDSs) is to maximize detection accuracy as well as minimize corresponding costs. In this paper, I present a cost-based approach utilizing decision tree and neural network models with GPU acceleration for fast intrusion detection. In this approach, both computational cost and time cost are considered, and different models are trained with respect to the types of protocols and services. Empirical experiments are carried out on off-line benchmark dataset NSL-KDD.

@@

\section{Introduction}
With the development of computing and network technology, Internet has became an important part of our daily life. However, this popularity has also brought many security issues. 
Intrusion Detection Systems(IDSs) are designed to detect attacks. Specifically, it helps to discover, determine, and identify unauthorized use, duplication, alteration, and destruction of information systems\cite{mukkamala2005cyber}. Additionally, IDSs are able to monitor the system and alert network administrators if necessary.

With respect to the types of detectable attacks, there are mainly three types of IDSs: misuse-based, anomaly-based, and hybrid\cite{buczak2016survey} IDS.
Misuse-based IDSs, which are also called signature-based IDSs are designed to detect known attacks by comparing new packets and the predefined signatures stored in the database.
Anomaly-based IDSs aims at modeling the normal network, and anomalies can be detected when happened. Hybrid IDSs combine the former two techniques together.

Also with respect to the using environment, IDSs can be divided as the ones used in real-time environment or used in off-line environment. Off-line IDSs have been studied extensively recently especially based on data mining and machine learning techniques, such as K-means\cite{lin2015cann} and ANN\cite{ingre2015performance} etc. And excellent detection results with high accuracies and low false alarm rates on the off-line datasets are accomplished.

However, these IDSs always fail to provide real-time detection, which means they can only detect intrusions after the they already happened for a while and are not capable in defending the host from an attack at all. This is because the excellent off-line performance is often achieved thanks to the comprehensive extracted features and long time complicated computing. So a balance need to be maintained between accuracy and efficiency in real-time IDSs.

In this paper, the aim is to develop an IDS which can be utilized in real-time environment with high detection accuracy and low cost. For the minimization of computational cost, I present a cost-based approach for fast intrusion detection. The traditional machine learning models are replaced by neural networks, which enable the utilizing of GPU acceleration. 
Also, the time cost is considered, and according to the time cost theory in \cite{lee2001real}, low level features are utilized with high priority. The benchmark dataset NSL-KDD is used as testbed for evaluation, which is a new version of dataset KDD Cup 1999 dataset\cite{cup2007available} and has significant improvements compared with KDD Cup 1999 dataset\cite{tavallaee2009detailed}.

@@

\begin{itemize}
\item \textbf{Denial of Service Attack (DoS):} This type of attacks aim at making the services of target computer unavailable to its legitimate users by increasing the number of significant resource demanding processes to exhaust the target's network bandwidth , CPU load, or memory etc\cite{bhuyan2014detecting}. However, with the development of detection and IP tracking techniques, old single source attacks can be easily detected and the source can be rebuffed or shut down. Many kinds of  distributed Dos(DDos) attack techniques have been developed. One of the effective approaches to detect DDos attacks is to capture and recognize patterns of the DDos affected traffic\cite{robinson2015ranking}. This survey\cite{zargar2013survey} is recommended for more information.
\item \textbf{User to Root Attack (U2R)} This type of attacks refer to the behavior that the attacker tries to gain root access of the system. Usually the attacker enters the system as a normal user first and try to exploit some vulnerabilities in the system to gain illegal root access.
\item \textbf{Remote to Local Attack (R2L):} This type of attacks occur when an attacker tries to gain local access as a legitimate user on a remote server where actually he or she has no legitimate accounts. R2L attacks are often carried out by sending packets over network to the remote server\cite{mukkamala2005intrusion}. Usually, this illegal accesses are gained by taking advantages of vulnerabilities on the server.
\item \textbf{Probing Attack:} This type of attacks are basically an information collection behavior. The attacker scans the computers in a specific network to gather information or to exploit vulnerabilities.
For more information about IDSs specially designed for probing attacks, this work \cite{zargar2009identification} is recommended.
\end{itemize}

@@

\subsubsection{Misuse-based Intrusion Detection}
The misuse-based intrusion detection is also known as signature-based or knowledge-based intrusion detection. A signature is a pattern or string that corresponds to a known attack or threat\cite{liao2013intrusion}. So in the misuse-based IDS, the detection procedure for new coming packets is to compare them with the known signatures for recognizing possible intrusions.

The main advantage of misuse-based methods is that the method is effective for detecting known type of attacks and the detection results are free from false alarms. However, there are also some disadvantages. Firstly, it's usually hard to keep the signature database up to date. Secondly, this type of IDSs are incapable in detecting zero-day attacks or any attacks whose signatures are not included in the database, so their abilities are quite limited.

\subsubsection{Anomaly-based Intrusion Detection}
An anomaly is a deviation from a known behavior\cite{liao2013intrusion}. The anomaly-based techniques aim at modeling normal status including regular activities, normal network connections, hosts or users etc. And new packets different from normal status can be detected, the anomalies are considered to be an attack.

@@

\subsection{Feature Selection}\label{sec:feature-selection}
The reduction of features and feature selection are very important issues in increasing the efficiency of the detection.

This work\cite{chebrolu2005feature} aims at reducing the features used for misuse-based techniques, since the comparison between the whole packet and all the signatures can be time demanding and actually redundant. Also, useful feature selection algorithms are proposed in the paper.
Furthermore, the features on one specific dataset KDD Cup 1999 dataset are analyzed by comparing the relevance between each features and information gain is employed to find the most discriminating features for each class\cite{kayacik2005selecting}.

Unsupervised machine learning techniques such as PCA\cite{liu2007hierarchical}\cite{lakhina2010feature} and K-means\cite{kumar2013k} are employed for feature selection process. And all of these methods demonstrates good performance on reducing the dimension of features and maintaining only the most representative ones.

However, as these feature selection methods are nearly all used in off-line detection, they ignore the time cost brought by the feature extraction process. So the selected features may bring severe time delay. In the cost-based approach presented in this paper, the cost factors are considerer, and a balance between cost and accuracy is maintained.

\subsection{Machine Learning Approaches}
In this section, I present a brief introduce of the utilization of machine learning techniques in intrusion detection. Specific machine learning algorithms are beyond the scope of this paper, and this book\cite{bishop2006pattern} is highly recommended for more knowledge on pattern recognition machine learning.

\subsubsection{Supervised Learning}
Supervised learning can be divided into generative models and deterministic models, and both classes contain plenty of proposed models. Some of the popular supervised learning methods are listed as follows: Artificial Neural Network, Bayesian Network, Decision Tree, Genetic Algorithms, Hidden Markov Models, Naive Bayes, Support Vector Machine etc.

Nearly all the mentioned techniques in supervised learning are used in intrusion detection, only a few papers are listed here\cite{ingre2015performance}\cite{sommer2010outside}\cite{qiao2002anomaly}. Comprehensive computational complexity analysis and concrete introduction of the supervised machine learning techniques can be found in \cite{buczak2016survey} and \cite{tsai2009intrusion}.

\subsubsection{Unsupervised Learning}
One of the most frequently used unsupervised learning techniques is \textbf{clustering}.  
This method aims at clustering the data with some similar properties together. The main advantage of clustering is that it makes the labeling work not necessary. Specifically, for intrusion detection task, this type of IDS can learn from audit data without labeling or requiring the network administrator to provide concrete attacks' information.

There are many clustering methods that demonstrate excellent performance in the evaluation\cite{hendry2008intrusion}\cite{lin2015cann}\cite{kumar2013k}\cite{tsai2010triangle}. K-means, K-NN are the most frequently used unsupervised learning approaches. Also, some work combine K-means and K-NN together and better results are achieved\cite{lin2015cann}. However, as mentioned in Sec \ref{sec:feature-selection}, feature selection and reduction are urgently needed in methods like K-NN, because computing the similarities between high dimensional features are really expensive.

\subsubsection{Ensemble Learning}
Actually, the ensemble learning approaches do not belong to supervised or unsupervised learning.
Basically, the idea of ensemble learning is to use multiple weak models to build a strong model. This is because trying to solve some complicated tasks with a single model may lead the model to be extremely complex. 
And it is known that complex models with lots of parameters usually tend to over fit on the training data.
Bagging and boosting are frequently used approaches in ensemble learning, this work\cite{dietterich2000experimental} gives a good example for utilizing ensemble learning methods in decision tree models.

@@

Ensemble learning methods are extensively utilized in intrusion detection\cite{zhang2008random}\cite{mukkamala2005intrusion}. The type of single models used for the aggregation can be nearly all kinds of previous mentioned machine learning models.

@@

In the approach proposed in this paper, both the supervised learning and ensemble learning methods are employed. The combination of decision tree model and neural networks model demonstrates a great improvement compared with using single model for prediction in NSL-KDD dataset. For example, random forest is used both in mis-use based IDS\cite{gharibian2007comparative} and anomaly-based IDS\cite{zhang2008random}.

@@

\subsubsection{DARPA 1998 and DARPA 1999 Datasets}
One of the most commonly used datasets are the Defense Advanced Research Projects Agency (DARPA)
1998 and DARPA 1999 datasets\cite{lippmann2000evaluating}\cite{lippmann20001999}. Both of the datasets are created by the Cyber Systems and Technology Group of the Massachusetts Institute of Technology Lincoln Laboratory (MIT/LL). The datasets contain both network and operating system data. Totally, seven weeks of training data and two weeks of test data are collected, and attack simulations are organized during both of the periods. Notice that for anomaly detection, there are some types of attacks only appear in the test data.

@@

\subsubsection{NSL-KDD Dataset}
As suggested in \cite{tavallaee2009detailed}, the original KDD Cup 1999 dataset has two important issues which highly affects the performance of evaluated systems, and results in a very poor evaluation of anomaly detection approaches.The NSL-KDD dataset consists of only the selected records of the complete KDD Cup 1999 dataset and free from the mentioned shortcomings in \cite{tavallaee2009detailed}.

@@

Neural network models are at the leaf nodes for the detection tree, they are trained for each protocol type and service respectively. Actually due to the less complexity for packets detection, neural network models perform nearly the same as traditional classifiers. The main reason of utilizing neural network models in this paper is the excellent GPU acceleration support for deep learning libraries, which will reduce the computational cost to some extent.

@@

Maybe you could notice that in the SVM method, the computational cost may even decay as the number of feature dimensions increase. This is because in the SVM model, only the training data on the decision boundary is used in the calculation for prediction, so the actual computational cost does not only depend on the dimension of features. For more theoretical knowledge of SVM, this book\cite{bishop2006pattern} is recommended.

>>>>

The aim of real time Intrusion Detection Systems(IDSs) is to maximize detection accuracy as well as minimize corresponding costs. In this paper, I present a cost-based approach utilizing decision tree model and neural network models with GPU acceleration for fast intrusion detection. In this approach, both computational cost and time cost are considered, and different models are trained with respect to the types of protocols and services. Empirical experiments are carried out on off-line benchmark dataset NSL-KDD.

@@

\section{Introduction}
With the development of computing and network technology, Internet has became an important part of our daily life. However, this popularity has also brought many security issues. 
Intrusion Detection Systems(IDSs) are designed to detect attacks. Specifically, they help to discover, determine, and identify unauthorized use, duplication, alteration, and destruction of information systems\cite{mukkamala2005cyber}. Additionally, IDSs are able to monitor the system and alert network administrators if necessary.

With respect to the types of detectable attacks, there are mainly three types of IDSs: misuse-based, anomaly-based, and hybrid\cite{buczak2016survey} IDSs.
Misuse-based IDSs, which are also called signature-based IDSs, are designed to detect known attacks by comparing new packets and the predefined signatures stored in the database.
Anomaly-based IDSs aims at modeling the normal network, and anomalies can be detected when happened. Hybrid IDSs combine the former two techniques together.

Also with respect to the using environment, IDSs can be divided as the ones used in real-time environment or used in off-line environment. Off-line IDSs have been studied extensively recently especially based on data mining and machine learning techniques, such as K-means\cite{lin2015cann} and ANN\cite{ingre2015performance}. And excellent detection results with high accuracies and low false alarm rates on the off-line datasets are accomplished.

However, these IDSs always fail to provide real-time detection, which means they can only detect intrusions after the they already happened for a while and are not capable of defending the host from attacks at all. 
This is because the excellent off-line performance is often achieved thanks to the comprehensive extracted features and long time complicated computing, which are impossible to get in real-time environment. So a balance need to be maintained between accuracy and efficiency in real-time IDSs.

In this paper, the aim is to develop an IDS which can be utilized in real-time environment with high detection accuracy and low costs. For the minimization of computational cost, the traditional machine learning models are replaced by neural network models, which enable the utilizing of GPU acceleration. 
Also, the time cost is considered, and similar with the time cost theory in \cite{lee2001real}, low level features are utilized with high priority. The benchmark dataset NSL-KDD is used as testbed for evaluation, which is a new version of KDD Cup 1999 dataset\cite{cup2007available} and significant improvement has been made compared with original KDD Cup 1999 dataset\cite{tavallaee2009detailed}.

@@

\begin{itemize}
\item \textbf{Denial of Service Attack (DoS):} This type of attacks aim at making the services of target computer unavailable to its legitimate users by increasing the number of resource demanding processes to exhaust the target's network bandwidth , CPU load, or memory etc\cite{bhuyan2014detecting}. However, with the development of detection and IP tracking techniques, old single source attacks can be easily detected and the source can be rebuffed or shut down. In recent years, many kinds of distributed Dos(DDos) attack techniques have been developed. One of the effective approaches to detect DDos attacks is to capture and recognize patterns of the DDos affected traffic\cite{robinson2015ranking}. This survey\cite{zargar2013survey} is recommended for more information.
\item \textbf{User to Root Attack (U2R)} This type of attacks refer to the behavior that the attacker tries to gain root access of the system. Usually the attacker enters the system as a normal user first and tries to exploit some vulnerabilities in the system to gain illegal root access.
\item \textbf{Remote to Local Attack (R2L):} This type of attacks occur when an attacker tries to gain local access as a legitimate user on a remote server where actually he or she has no legitimate accounts. R2L attacks are often carried out by sending packets over network to the remote server\cite{mukkamala2005intrusion}. Usually, this illegal accesses are gained by taking advantages of vulnerabilities on the server.
\item \textbf{Probing Attack:} This type of attacks are basically an information collection behavior. The attacker scans the computers in a specific network to gather information or to exploit vulnerabilities.
For more information about IDSs specially designed for probing attacks, this work \cite{zargar2009identification} is recommended.
\end{itemize}

@@

\subsubsection{Misuse-based Intrusion Detection}
The misuse-based intrusion detection is also known as signature-based or knowledge-based intrusion detection. A signature is a pattern or string that corresponds to a known attack or threat\cite{liao2013intrusion}. So in the misuse-based IDSs, the detection procedure for new coming packets is to compare them with the known signatures to recognize possible intrusions.

The main advantage of misuse-based methods is that the method is effective for detecting known types of attacks and the detection results are free from false alarms. However, there are also some disadvantages. Firstly, it's usually hard to keep the signature database up to date. Secondly, this type of IDSs are incapable in detecting zero-day attacks or any attacks whose signatures are not included in the database, so their abilities are quite limited.

\subsubsection{Anomaly-based Intrusion Detection}
An anomaly is a deviation from a known behavior\cite{liao2013intrusion}. The anomaly-based IDSs aim at modeling normal status including regular activities, normal network connections, hosts and users etc. And new packets different from normal status can be detected, the anomalies are considered to be an attack.

@@

\subsection{Feature Selection}\label{sec:feature-selection}
The reduction of features and feature selection are very important issues in increasing the efficiency of intrusion detection.

This work\cite{chebrolu2005feature} aims at reducing the features used for misuse-based IDSs, since the comparison between the whole packet and all the signatures can be time demanding and actually redundant. Also, useful feature selection algorithms are proposed in the paper.
Furthermore, the features on one specific dataset KDD Cup 1999 dataset are analyzed by comparing the relevance between each features and information gain is employed to find the most discriminating features for each class\cite{kayacik2005selecting}.

Unsupervised machine learning techniques such as PCA\cite{liu2007hierarchical}\cite{lakhina2010feature} and K-means\cite{kumar2013k} are employed for feature selection process. And all of these methods demonstrate good performance on reducing the dimension of features and reserving only the most representative ones.

However, since nearly all of these feature selection methods are used in off-line environment, they ignore the time cost brought by the feature extraction process. So the proposed selected features may bring severe time delay. In the cost-based approach presented in this paper, the time cost factors are considerer, and a balance between efficiency and accuracy is maintained.

\subsection{Machine Learning Approaches}
In this section, I present a brief introduce of the utilization of machine learning techniques in intrusion detection. Specific machine learning algorithms are beyond the scope of this paper, and this book\cite{bishop2006pattern} is highly recommended for more knowledge on pattern recognition and machine learning.

\subsubsection{Supervised Learning}
Supervised learning can be further divided into two classes: generative models and deterministic models. Both classes contain plenty of proposed models. Some of the popular supervised learning methods are listed as follows: Artificial Neural Network, Bayesian Network, Decision Tree, Genetic Algorithms, Hidden Markov Models, Naive Bayes, Support Vector Machine etc.

Nearly all the mentioned techniques in supervised learning are used in intrusion detection, only a few papers are listed here\cite{ingre2015performance}\cite{sommer2010outside}\cite{qiao2002anomaly}. Comprehensive computational complexity analysis and concrete introduction results can be found in \cite{buczak2016survey} and \cite{tsai2009intrusion}.

\subsubsection{Unsupervised Learning}
One of the most frequently used unsupervised learning techniques is \textbf{clustering}.  
This method aims at clustering the data with similar properties together. The main advantage of clustering is that it makes the labeling work not necessary. Specifically, for intrusion detection task, this type of IDS can learn from audit data without labeling or requiring the network administrator to provide concrete information about the attacks.

There are many clustering methods that demonstrate excellent performance during evaluation\cite{hendry2008intrusion}\cite{lin2015cann}\cite{kumar2013k}\cite{tsai2010triangle}. K-means, K-NN are the most frequently used unsupervised learning approaches. Also, some work combine K-means and K-NN together and better results are achieved\cite{lin2015cann}. However, as mentioned in Sec \ref{sec:feature-selection}, feature selection and reduction are urgently needed in methods like K-NN, because computing the similarities between high dimensional features are really expensive.

\subsubsection{Ensemble Learning}
Actually, the ensemble learning approaches do not belong to supervised or unsupervised learning.
Basically, the idea of ensemble learning is to use multiple weak models to build a strong model. This is because trying to solve a complicated task with a single model may lead the model to be extremely complex. 
And it is known that complex models with lots of parameters often tend to over fit on the training data. 
The aggregation of many weak models can avoid over fitting to some extent.
Bagging and boosting are frequently used approaches in ensemble learning, this work\cite{dietterich2000experimental} gives a good example for utilizing ensemble learning methods in decision tree models.

@@

Ensemble learning methods are extensively utilized in intrusion detection\cite{zhang2008random}\cite{mukkamala2005intrusion}. The type of single models used for the aggregation can be nearly all kinds of previous mentioned machine learning models. For example, random forest is used both in mis-use based IDS\cite{gharibian2007comparative} and anomaly-based IDS\cite{zhang2008random}.

@@

In the approach proposed in this paper, both the supervised learning and ensemble learning methods are employed. The combination of decision tree model and neural networks models demonstrates a great improvement compared with using single model for prediction on NSL-KDD dataset.

@@

\subsubsection{DARPA 1998 and DARPA 1999 Datasets}
One of the most commonly used datasets are the Defense Advanced Research Projects Agency (DARPA)
1998 and DARPA 1999 datasets\cite{lippmann2000evaluating}\cite{lippmann20001999}. Both of the datasets are created by the Cyber Systems and Technology Group of the Massachusetts Institute of Technology Lincoln Laboratory (MIT/LL). The datasets contain both network and operating system data. Totally, seven weeks of training data and two weeks of test data are collected, and attack simulations are organized during both of the periods. Notice that for anomaly detection, there are some types of attacks only appear in the testing data.

@@

\subsubsection{NSL-KDD Dataset}
As suggested in \cite{tavallaee2009detailed}, the original KDD Cup 1999 dataset has two important shortcomings which highly affect the performance of evaluated systems, and result in a very poor evaluation of anomaly detection approaches.The NSL-KDD dataset consists of only the selected records of the complete KDD Cup 1999 dataset and free from the mentioned issues in \cite{tavallaee2009detailed}.

@@

Neural network models are the leaf nodes for the decision tree, they are trained for each protocol type and service respectively. Actually due to the less complexity for packets detection, neural network models perform nearly the same as traditional classifiers. The main reason of utilizing neural network models in this paper is the excellent GPU acceleration support for deep learning libraries, which will reduce the computational cost to some extent.

@@

Maybe you could notice that in the SVM method, the computational cost may even decay as the number of feature dimension increase. This is because in the SVM model, only the training data on the decision boundary is used in the calculation for prediction, so the actual computational cost does not only depend on the dimension of features. For more theoretical knowledge of SVM, this book\cite{bishop2006pattern} is recommended.
